---
title: "Associations (Kruskal-Wallis) with questionnaire variables"
author: "Leo Dahl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
```

# Description

> Looking at associations between proteins and questionnaire variables.

# Packages and data

```{r packages, message=F, warning=F}

rm(list = ls())
options(stringsAsFactors = F)
library(tidyverse)
library(DT)
library(openxlsx)
library(ordinal) # For clm for ordinal logistic regression
library(ggvenn)
library(patchwork)

```

***

Data preparation
================

Load files
----------

```{r}

source("load_cov_npx.R")

cov_npx <- load_cov_npx("abspqn_excl")

source("sep_datasets.R")
out_dir <- "../results/association_tests_lm_kw/"
```

Loaded data set **`r cov_npx$data_set`**  
from **`r cov_npx$data_path`**, with the following treatment:  
`r cov_npx$norm`


```{r}

npx <- cov_npx$npx
sinfo <- cov_npx$sinfo
binder <- cov_npx$binder

# Turn into character to not get unexpected behaviour from factor
sinfo$grp <- as.character(sinfo$grp)

# Merge duplicates/triplicates
sep_dat <- sep_datasets(npx, sinfo)
npx_umap <- sep_dat$npx_umap
npx_spk <- sep_dat$npx_spk
sinfo_umap <- sep_dat$sinfo_umap %>% mutate(serostatus = case_when(serostatus == "no reactivity" ~ "Seronegative, IgM-IgG-",
                                                                   serostatus == "IgG + IgM" ~ "Seropositive, IgM+IgG+"))
sinfo_spk <- sep_dat$sinfo_spk %>% mutate(serostatus = case_when(serostatus == "IgM" ~ "Early phase, IgM+IgG-",
                                                                 serostatus == "IgG" ~ "Late phase, IgM-IgG+"))
rm(sep_dat)

```

Also get studyset 3

```{r}

ds3 <- load_cov_npx(paste0("ds3_", cov_npx$data_set)) %>% merge_repl()

sinfo_ds3 <- ds3$sinfo %>% rename(serostatus = grp) %>%
  mutate(serostatus = case_when(serostatus == "Negative" ~ "Seronegative, IgG-",
                                serostatus == "Positive" ~ "Seropositive, IgG+"))
npx_ds3 <- ds3$npx %>% filter(sample_id %in% sinfo_ds3$sample_id)
binder_ds3 <- ds3$binfo

```

Loaded data set from **`r ds3$data_path`**

***

# Performing logistic regression

```{r}
# Put all NPX values in one dataframe, add panel info and sample annotation
npx_in <- rbind(
  # Studyset1
  npx_umap %>% as.data.frame() %>% rownames_to_column("sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    mutate(studyset = "Studyset1") %>%
    left_join(binder %>% select(unique_id, panel_name), by = c("assay" = "unique_id")) %>%
    left_join(sinfo %>% select(sample_id, Age_grp, Sex), by = "sample_id") %>%
    rename(panel = panel_name, age_group = Age_grp, sex = Sex) %>%
    as.data.frame(),
  # Studyset2
  npx_spk %>% as.data.frame() %>% rownames_to_column("sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    mutate(studyset = "Studyset2") %>%
    left_join(binder %>% select(unique_id, panel_name), by = c("assay" = "unique_id")) %>%
    left_join(sinfo %>% select(sample_id, Age_grp, Sex), by = "sample_id") %>%
    rename(panel = panel_name, age_group = Age_grp, sex = Sex) %>%
    as.data.frame(),
  # Studyset3
  npx_ds3 %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    mutate(studyset = "Studyset3") %>%
    left_join(ds3$binfo %>% select(assay, panel), by = c("assay")) %>%
    left_join(sinfo_ds3 %>% select(sample_id, age_group, sex), by = "sample_id") %>%
    as.data.frame()
) %>%
  # Convert variables to factors for regression, order age in ascending order
  mutate(age_group = factor(age_group, levels = sort(unique(age_group))),
         sex = factor(sex))

logreg <- npx_in %>%
  pivot_longer(cols = -c(sample_id, assay, npx, studyset, panel)) %>%
  # Logreg per studyset, assay, variable (age, sex)
  group_by(studyset, assay, name) %>% nest() %>%
  mutate(logreg_out = map(data, ~ {
    # Binomial logreg for sex, ordinal logreg for age group
    if (name == "sex") {
      logreg_obj <- glm(formula = value ~ npx, data = .x, family = binomial)
      test <- "logreg"
    } else if (name == "age_group") {
      logreg_obj <- clm(formula = value ~ npx, data = .x, link = "logit")
      test <- "ordinal logreg"
    }
    
    logreg_p <- summary(logreg_obj)$coefficients["npx", "Pr(>|z|)"]
    logreg_est <- summary(logreg_obj)$coefficients["npx", "Estimate"]
    
    return(data.frame("assay" = assay, "studyset" = studyset, "variable" = name, "test" = test, "estimate" = logreg_est, "pval" = logreg_p))
  })) %>%
  pull(logreg_out) %>%
  data.table::rbindlist() %>%
  # Compute FDR per studyset and variable
  ungroup() %>% group_by(studyset, variable) %>%
  mutate(fdr = p.adjust(pval, method = "fdr")) %>%
  ungroup()

```

## Beta plots

```{r, fig.height=4, fig.width=11}

# p < 0.05
beta_plots_p <- lapply(names(venn_dat_p), function(v) {
  # Compare betas for overlapping proteins, two studysets at a time
  combs <- combn(names(venn_dat_p[[v]]), 2, simplify = F)
  
  lapply(combs, function(cmb) {
    overlap_prot <- Reduce(intersect, venn_dat_p[[v]][cmb])

    # Make plot
    logreg %>% filter(studyset %in% cmb, assay %in% overlap_prot, variable == v) %>%
      select(studyset, variable, assay, estimate) %>%
      pivot_wider(id_cols = c(variable, assay), names_from = studyset, values_from = estimate) %>%
      ggplot(aes(x = .data[[cmb[1]]], y = .data[[cmb[2]]])) +
      geom_point() +
      # Mark quadrants to easily see how well they agree
      geom_hline(yintercept = 0, linetype = "dashed") + geom_vline(xintercept = 0, linetype = "dashed") +
      labs(x = paste0("Beta ", cmb[1]), y = paste0("Beta ", cmb[2])) +
      theme_classic()
    
  }) %>%
    wrap_plots(nrow = 1) +
    plot_annotation(title = paste0("Overlapping ", v, " associations, p < 0.05"))
  
})

# fdr < 0.05
beta_plots_fdr <- lapply(names(venn_dat_fdr), function(v) {
  # Compare betas for overlapping proteins, two studysets at a time
  combs <- combn(names(venn_dat_fdr[[v]]), 2, simplify = F)
  
  lapply(combs, function(cmb) {
    overlap_prot <- Reduce(intersect, venn_dat_fdr[[v]][cmb])

    # Make plot
    logreg %>% filter(studyset %in% cmb, assay %in% overlap_prot, variable == v) %>%
      select(studyset, variable, assay, estimate) %>%
      pivot_wider(id_cols = c(variable, assay), names_from = studyset, values_from = estimate) %>%
      ggplot(aes(x = .data[[cmb[1]]], y = .data[[cmb[2]]])) +
      geom_point() +
      # Mark quadrants to easily see how well they agree
      geom_hline(yintercept = 0, linetype = "dashed") + geom_vline(xintercept = 0, linetype = "dashed") +
      labs(x = paste0("Beta ", cmb[1]), y = paste0("Beta ", cmb[2])) +
      theme_classic()
    
  }) %>%
    wrap_plots(nrow = 1) +
    plot_annotation(title = paste0("Overlapping ", v, " associations, FDR < 0.05"))
  
})

beta_plots_p
beta_plots_fdr

```

Listed below are the proteins with different betas in different studysets (pvalue < 0.05)

```{r}
beta_diff <- sapply(names(venn_dat_p), function(v) {
  # Compare betas for overlapping proteins, two studysets at a time
  combs <- combn(names(venn_dat_p[[v]]), 2, simplify = F)
  
  lapply(combs, function(cmb) {
    overlap_prot <- Reduce(intersect, venn_dat_p[[v]][cmb])

    # Get those with different beta signs
    diff_prot <- logreg %>% filter(studyset %in% cmb, assay %in% overlap_prot, variable == v) %>%
      select(studyset, variable, assay, estimate) %>%
      mutate(beta_sign = sign(estimate)) %>%
      group_by(assay) %>%
      filter(beta_sign[1] != beta_sign[2]) %>%
      arrange(assay)
  })
  
}, simplify = F)

beta_diff
```


# Logistic regression for association between proteins and serostatus

## Load adjusted data

```{r}

source("load_cov_npx.R")

cov_npx <- load_cov_npx("abspqn_adj")

source("sep_datasets.R")
```

Loaded data set **`r cov_npx$data_set`**  
from **`r cov_npx$data_path`**, with the following treatment:  
`r cov_npx$norm`


```{r}

npx <- cov_npx$npx
sinfo <- cov_npx$sinfo
binder <- cov_npx$binder

# Turn into character to not get unexpected behaviour from factor
sinfo$grp <- as.character(sinfo$grp)

# Merge duplicates/triplicates
sep_dat <- sep_datasets(npx, sinfo)
npx_umap <- sep_dat$npx_umap
npx_spk <- sep_dat$npx_spk
sinfo_umap <- sep_dat$sinfo_umap %>% mutate(serostatus = case_when(serostatus == "no reactivity" ~ "Seronegative, IgM-IgG-",
                                                                   serostatus == "IgG + IgM" ~ "Seropositive, IgM+IgG+"))
sinfo_spk <- sep_dat$sinfo_spk %>% mutate(serostatus = case_when(serostatus == "IgM" ~ "Early phase, IgM+IgG-",
                                                                 serostatus == "IgG" ~ "Late phase, IgM-IgG+"))
rm(sep_dat)

```

Also get studyset 3

```{r}

ds3 <- load_cov_npx(paste0("ds3_", cov_npx$data_set)) %>% merge_repl()

sinfo_ds3 <- ds3$sinfo %>% rename(serostatus = grp) %>%
  mutate(serostatus = case_when(serostatus == "Negative" ~ "Seronegative, IgG-",
                                serostatus == "Positive" ~ "Seropositive, IgG+"))
npx_ds3 <- ds3$npx %>% filter(sample_id %in% sinfo_ds3$sample_id)
binder_ds3 <- ds3$binfo

```

Loaded data set from **`r ds3$data_path`**

## Regression

```{r}
# Put all NPX values in one dataframe
npx_in <- rbind(
  # Studyset1
  npx_umap %>% as.data.frame() %>% rownames_to_column("sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    mutate(studyset = "Studyset1") %>%
    left_join(sinfo_umap %>% select(sample_id, serostatus), by = "sample_id") %>%
    as.data.frame(),
  # Studyset2
  npx_spk %>% as.data.frame() %>% rownames_to_column("sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    mutate(studyset = "Studyset2") %>%
    left_join(sinfo_spk %>% select(sample_id, serostatus), by = "sample_id") %>%
    as.data.frame(),
  # Studyset3
  npx_ds3 %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    mutate(studyset = "Studyset3") %>%
    left_join(sinfo_ds3 %>% select(sample_id, serostatus), by = "sample_id") %>%
    as.data.frame()
)

logreg_res <- npx_in %>%
  # Perform per studyset and protein
  group_by(studyset, assay) %>% nest() %>%
  mutate(test_res = map(data, ~ {
    
    logreg_obj <- glm(serostatus ~ npx, family = "binomial", data = .x %>% mutate(serostatus = factor(serostatus)))
    
    
    logreg_p <- summary(logreg_obj)$coefficients["npx", "Pr(>|z|)"]
    logreg_est <- summary(logreg_obj)$coefficients["npx", "Estimate"]
    
    return(data.frame("assay" = assay, "studyset" = studyset, "test" = "logreg", "estimate" = logreg_est, "pval" = logreg_p))
    
  })) %>%
  pull(test_res) %>%
  data.table::rbindlist() %>%
  ungroup() %>%
  group_by(studyset) %>%
  mutate(fdr = p.adjust(pval, method = "fdr")) %>%
  ungroup()

```

```{r}
datatable(logreg_res) %>% formatSignif(columns = 3:4, digits = 3)
```

Save tables

```{r}
tbl_out <- rbind(
  logreg,
  logreg_res %>% mutate(variable = "serostatus") %>% relocate(variable, .after = 2)
) %>%
  rename(unique_id = assay) %>%
  left_join(binder %>% select(assay, unique_id, gene_name, uniprot_ID), by = "unique_id") %>%
  relocate(assay, gene_name, uniprot_ID) %>%
  # Reorder binders in same order as in NPX data
  mutate(unique_id = factor(unique_id, levels = binder$unique_id)) %>%
  arrange(unique_id) %>%
  select(-unique_id, -test) %>%
  # Make wider to fit all in one table instead of three
  pivot_wider(id_cols = c(assay, gene_name, uniprot_ID, variable),
              names_from = studyset, values_from = c(estimate, pval, fdr))

# Get combined p-values
comb_p <- read.xlsx(paste0(out_dir, "Associations.xlsx")) %>%
  left_join(binder %>% select(assay, gene_name), by = "assay") %>%
  relocate(gene_name, .after = 1)

tbl_out <- list("Association logreg" = tbl_out, "Association combined pval" = comb_p)

write.xlsx(tbl_out, file = paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_assoc_test_results.xlsx"), asTable = T)
```


# Session information

```{r session information}
sessionInfo()
```
