---
title: "Binder IQR"
author: "Leo Dahl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
```

# Description

> Plotting binder IQR

# Packages and data

```{r packages, message=F, warning=F}

rm(list = ls())

library(tidyverse)
library(patchwork)
library(openxlsx)

```

Load data

```{r load data}

source("load_cov_npx.R")
cov_npx <- load_cov_npx("abspqn")

source("sep_datasets.R")
out_dir <- "../results/binder_iqr/"
```

```{r}

npx <- cov_npx$npx
sinfo <- cov_npx$sinfo
binder <- cov_npx$binder

# Turn into character to not get unexpected behaviour from factor
sinfo$grp <- as.character(sinfo$grp)

# Merge duplicates/triplicates
sep_dat <- sep_datasets(npx, sinfo)
npx_umap <- sep_dat$npx_umap
npx_spk <- sep_dat$npx_spk
sinfo_umap <- sep_dat$sinfo_umap
sinfo_spk <- sep_dat$sinfo_spk
rm(sep_dat)

```

Loaded data set **`r cov_npx$data_set`**  
from **`r cov_npx$data_path`**, with the following treatment:  
`r cov_npx$norm`

Load studyset 3 as well

```{r}
ds3 <- load_cov_npx(paste0("ds3_", cov_npx$data_set)) %>% merge_repl()
sinfo_ds3 <- ds3$sinfo %>% rename(serostatus = grp)
npx_ds3 <- ds3$npx %>% filter(sample_id %in% sinfo_ds3$sample_id)
binder_ds3 <- ds3$binfo
```

Studyset3 version: **`r ds3$data_set`** from **`r ds3$data_path`**

Finally, put all studysets into one list

```{r}
dat <- list("studyset1" = list("npx" = npx_umap,
                              "sinfo" = sinfo_umap,
                              "name" = "studyset1"),
            "studyset2" = list("npx" = npx_spk,
                              "sinfo" = sinfo_spk,
                              "name" = "studyset2"),
            "studyset3" = list("npx" = ds3$npx %>% column_to_rownames("sample_id") %>% as.matrix(),
                              "sinfo" = ds3$sinfo %>% filter(sample_id %in% ds3$npx$sample_id) %>% rename(serostatus = grp),
                              "name" = "studyset3"))
# Make a long format data
dat <- lapply(dat, function(d) {
  npx_long <- d$npx %>% as.data.frame() %>% rownames_to_column("sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "assay", values_to = "npx") %>%
    left_join(d$sinfo %>% select(sample_id, serostatus), by = "sample_id") %>%
    filter(!is.na(serostatus)) %>%
    mutate(studyset = d$name)
}) %>%
  data.table::rbindlist()
```

# Computing and plotting IQR

```{r, fig.width=9, fig.height=3}
binder_iqr <- dat %>%
  group_by(studyset, assay) %>%
  summarise(iqr = IQR(npx), .groups = "keep") %>%
  ungroup() %>%
  # Add panel names and protein names without panel
  mutate(panel = binder[match(assay, binder$unique_id), "panel_name", drop = T],
         protein = binder[match(assay, binder$unique_id), "assay", drop = T])

# Make plots in loop instead of facetting to be able to set axis label orders individually
iqr_plt <- binder_iqr %>% group_by(studyset) %>% nest() %>%
  mutate(plt = map(data, ~ {
    x_order <- .x %>% arrange(iqr) %>% pull(protein)
    
    .x %>% mutate(protein = factor(protein, levels = x_order)) %>%
      ggplot(aes(x = protein, y = iqr)) +
      geom_point() +
      labs(title = studyset, x = "Protein", y = ifelse(studyset == "studyset1", "IQR", "")) +
      scale_color_brewer(palette = "Set2") +
      ylim(0, 2.5) +
      theme_classic() +
      theme(axis.text.x = element_blank(),
            axis.ticks.x = element_blank(),
            plot.title = element_text(hjust = 0.5),
            axis.text = element_text(size = 12),
            axis.title = element_text(size = 12),
            axis.line = element_line(linewidth = 0.8),
            legend.title = element_text(size = 12),
            legend.text = element_text(size = 12))
  })) %>%
  pull(plt) %>%
  wrap_plots(nrow = 1, guides = "collect")

iqr_plt

```

# Ranks and correlations

The proteins are ranked by their IQR values and the ranks are compared between studysets.

```{r}

# Add rank of protein to the data and make a bit wider
binder_iqr_wide <- binder_iqr %>%
  group_by(studyset) %>%
  mutate(rank = rank(iqr)) %>%
  pivot_wider(id_cols = protein, names_from = studyset, values_from = c(iqr, rank)) %>%
  # Add protein information
  left_join(cov_npx$binder %>% select(assay, gene_name, uniprot_ID, unique_id), by = c("protein" = "assay")) %>%
  rename(assay = protein) %>% relocate(assay, gene_name, uniprot_ID) %>% select(-unique_id)

# Get the correlations of IQR values between studysets and their p-values
iqr_cor <- lapply(combn(sort(unique(binder_iqr$studyset)), 2, simplify = F), function(comb) {
  ct <- cor.test(binder_iqr %>% filter(studyset == comb[1]) %>% arrange(protein) %>% pull(iqr),
                 binder_iqr %>% filter(studyset == comb[2]) %>% arrange(protein) %>% pull(iqr))
  data.frame("comparison" = paste0(comb[1], "-", comb[2]),
             "correlation" = ct$estimate,
             "pval" = ct$p.value)
}) %>% data.table::rbindlist() %>% as.data.frame()

iqr_cor
```

```{r}
# Save tables and plots
ggsave(filename = paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_iqr_plots.pdf"),
       plot = iqr_plt, height = 3.5, width = 10.5)

write.xlsx(x = list("IQR" = binder_iqr_wide, "IQR correlations" = iqr_cor),
           file = paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_iqr_table.xlsx"),
           asTable = T)
```


# Session information

```{r session information}
sessionInfo()
```
