---
title: "DBS CV"
author: "Leo Dahl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    code_folding: hide
    toc_float:
      collapsed: false 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
```


<style type="text/css">

h1 { 
  font-size: 25px;
  margin-top: 1.5cm;
  margin-bottom: 0.5cm;
}
h2 { 
  font-size: 18px;
  margin-top: 1cm;
  margin-bottom: 0.5cm;
}

</style>

Description
===========

> Coefficient of Variation (CV) plots are made for bridge samples and replicates in the DBS olink assays. 

```{r prep, message=F, results="hide"}

rm(list=ls())
library(tidyverse)
library(patchwork)
library(DT)
library(openxlsx)

```

Power-transformed NPX values (2^NPX) are used for CV inspection.


```{r}

source("load_cov_npx.R")

cov_npx <- load_cov_npx("trimmed")

```

Loaded data set **`r cov_npx$data_set`**  
from **`r cov_npx$data_path`**, with the following treatment:  
`r cov_npx$norm`


```{r}

npx <- cov_npx$npx
sinfo <- cov_npx$sinfo
binder <- cov_npx$binder

# Turn into character to not get unexpected behaviour from factor
sinfo$grp <- as.character(sinfo$grp)

```

Also get dataset 3

```{r}
source("sep_datasets.R")
ds3 <- load_cov_npx(paste0("ds3_", cov_npx$data_set)) %>% merge_repl()

sinfo_ds3 <- ds3$sinfo %>% rename(serostatus = grp) %>%
  mutate(serostatus = case_when(serostatus == "Negative" ~ "Seronegative, IgG-",
                                serostatus == "Positive" ~ "Seropositive, IgG+"))
npx_ds3 <- ds3$npx %>% filter(sample_id %in% sinfo_ds3$sample_id)
binder_ds3 <- ds3$binfo

```

Loaded data set from **`r ds3$data_path`**


```{r}
out_dir <- "../results/cv/"
```



```{r prep data}

# Linearise NPX values
npx <- 2 ** cov_npx$npx
binder <- cov_npx$binder
sinfo <- cov_npx$sinfo

# Exclude triplicate that is detected as an outlier
# Relabel some samples as "Replicates" and one as "Population" (from "Longitudinal") to use for sorting
# (code excluded)

```



```{r define functions}

cv <- function(in_vec) abs( sd(in_vec) / mean(in_vec) )*100

# Computing and plotting CVs
cv_compute <- function(sample_info, npx_mat, grp_val) {
  # sample_info, sample info data frame from which to pick samples
  # npx_mat, NPX matrix
  # grp, value to use for grouping column
  
  # Compute CVs
  cvs <- sample_info %>%
    filter(grp == grp_val) %>%
    select(unique_participant_ID, sample_id) %>%
    # Add NPX values
    left_join(npx_mat %>% as.data.frame() %>% rownames_to_column("sample_id"),
              by = "sample_id") %>%
    select(-sample_id) %>%
    pivot_longer(cols = -unique_participant_ID, names_to = "protein", values_to = "mfi") %>%
    group_by(unique_participant_ID, protein) %>%
    # Compute CVs per individual and per protein
    summarise(cv = cv(mfi), .groups = "keep") %>%
    ungroup()
  
  # Plot per individual
  plt_per_ind <- cvs %>%
    group_by(unique_participant_ID) %>%
    nest() %>%
    ungroup() %>%
    # Give the samples new names
    mutate(sample_name = paste0("Individual ", 1:nrow(.))) %>%
    group_by(sample_name) %>%
    mutate(plt = map(data, ~ {
      # Plot protein in CV order
      .x %>%
        arrange(cv) %>%
        mutate(protein = factor(protein, levels = protein)) %>%
        ggplot(aes(x = protein, y = cv)) +
        geom_point() +
        labs(title = sample_name, x = "Proteins, CV order", y = "CV%") +
        theme_classic(18) +
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank())
    })) %>%
    pull(plt)
  
  # Plot with individuals merged
  plt_merge <- cvs %>%
    group_by(protein) %>%
    summarise(median_cv = median(cv), .groups = "keep") %>%
    ungroup() %>%
    # Plot in CV order
    arrange(median_cv) %>%
    mutate(protein = factor(protein, levels = protein)) %>%
    ggplot(aes(x = protein, y = median_cv)) +
    geom_point() +
    labs(x = "Protein, CV order", y = "CV%") +
    theme_classic(18) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())

  # Median CV
  tbl_median <- cvs %>%
    group_by(protein) %>%
    summarise(median_cv = median(cv), .groups = "keep") %>%
    ungroup()
  
  # Table with CV for individuals and median CVs
  tbl_merge <- cvs %>%
    # Add new sample names and use as columns
    mutate(individual = paste0("Individual ", as.numeric(as.factor(unique_participant_ID)))) %>%
    pivot_wider(id_cols = protein, names_from = individual, values_from = cv) %>%
    # Add median CV
    left_join(tbl_median %>% rename("Median CV%" = "median_cv"), by = "protein") %>%
    rename("Binder" = "protein") %>%
    # Add more binder info
    left_join(binder %>% select(unique_id, gene_name, panel_name, uniprot_ID), by = c("Binder" = "unique_id")) %>%
    relocate(Binder, gene_name, panel_name, uniprot_ID) %>%
    rename("Gene name" = "gene_name", "Panel" = "panel_name", "UniProt ID" = "uniprot_ID") %>%
    # Sort in ascending CV order
    arrange(`Median CV%`)
  
  return(list("cvs" = cvs, "cv_median" = tbl_median, "cv_table" = tbl_merge,
              "plot_per_ind" = plt_per_ind, "plot_merge" = plt_merge))
}

```

Bridge samples
==============

```{r plot bridge cvs, fig.dim=c(12,10)}
# Inter-plate duplicates CV
inter_cv <- cv_compute(sinfo, npx, "Bridge")

wrap_plots(inter_cv$plot_per_ind) + plot_annotation(title = "Interplate CV%", theme = theme(plot.title = element_text(size = 20)))
```


```{r plot bridge cvs 2}
inter_cv$plot_merge + labs(title = "Median interplate CV%")
```

Triplicate samples
==================

```{r triplicate cv, fig.dim=c(10,5)}

intra_cv <- cv_compute(sinfo, npx, "Replicates")

wrap_plots(intra_cv$plot_per_ind) + plot_annotation(title = "Intraplate CV%", theme = theme(plot.title = element_text(size = 20)))
intra_cv$plot_merge + labs(title = "Median intraplate CV%")

```

Look at the normalised MFI values of the triplicates

```{r triplicate mfi}

tripl_mfi <- npx %>%
    as.data.frame() %>%
    rownames_to_column("sample_id") %>%
    left_join(sinfo %>% select(sample_id, grp, unique_participant_ID),
              by = "sample_id") %>%
    filter(grp == "Replicates") %>%
    select(-grp, -unique_participant_ID) %>%
    separate(col = sample_id, into = c("participant_id", "repl"), sep = "_") %>%
    pivot_longer(cols = -c("participant_id", "repl")) %>%
    ggplot(aes(x = name, y = value, colour = repl)) +
    geom_point() +
    scale_colour_brewer(palette = "Set2") +
    facet_wrap(~ participant_id, ncol = 1) +
    labs(x = "Protein", y = "Normalised MFI", colour = "Replicate") +
    theme_classic(16) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())

tripl_mfi

```

```{r}
cv_plt_fn <- paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_cv_plots.pdf")
# Wrap plots and add tags before saving
intra_plt <- wrap_plots(intra_cv$plot_per_ind)
intra_plt$patches$plots[[1]] <- intra_plt$patches$plots[[1]] + labs(tag = "A")
inter_plt <- wrap_plots(inter_cv$plot_per_ind)
inter_plt$patches$plots[[1]] <- inter_plt$patches$plots[[1]] + labs(tag = "B")

ggsave(filename = cv_plt_fn,
       plot = intra_plt / inter_plt + plot_layout(heights = c(1, 2)) & theme(plot.tag = element_text(size = 24)),
       height = 14, width = 10)
```

Plots of CVs of each individual saved as PDF file at `r `

CV tables
=========

```{r cv tables}
datatable(intra_cv$cv_table, caption = "Intraplate CV%, from triplicates")
datatable(inter_cv$cv_table, caption = "Interplate CV%, from bridging samples (duplicates)")
```


```{r save tables}
# Save in excel file
cv_tbl_out <- intra_cv$cv_table %>% select(-contains("Individual")) %>%
  rename("median_intraplate_cv" = "Median CV%") %>%
  left_join(inter_cv$cv_table %>% select(-contains("Individual")) %>%
              rename("median_interplate_cv" = "Median CV%"),
            by = c("Binder", "Gene name", "Panel", "UniProt ID")) %>%
  left_join(binder %>% select(assay, unique_id), by = c("Binder" = "unique_id")) %>%
  rename("gene_name" = "Gene name", "uniprot_ID" = "UniProt ID") %>% select(-Binder, -Panel) %>%
  relocate(assay, gene_name, uniprot_ID)

write.xlsx(cv_tbl_out, paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_cv_table.xlsx"))

```


Session information
===================

```{r session info}
sessionInfo()
```


