---
title: "Global overview of the three datasets"
author: "Leo Dahl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
```

# Description

> Provides a global overview of the datasets with PCA, UMAP, and median vs IQR plots. The last plot also classifies outliers by marking samples that fall outside of mean + 3SD thresholds of median NPX or IQR.

# Packages and data

```{r packages, message=F, warning=F}

rm(list = ls())
options(stringsAsFactors = F)
library(tidyverse)
library(ggrepel)
library(ggh4x)
library(umap)
library(patchwork)

```

```{r load data}

source("load_cov_npx.R")
cov_npx <- load_cov_npx("abspqn_excl")

out_dir <- "../results/global_overview/"
```

Loaded data set **`r cov_npx$data_set`**  
from **`r cov_npx$data_path`**, with the following treatment:  
`r cov_npx$norm`


```{r}

npx <- cov_npx$npx
sinfo <- cov_npx$sinfo
binder <- cov_npx$binder

# Turn into character to not get unexpected behaviour from factor
sinfo$grp <- as.character(sinfo$grp)

# Separate datasets, use complete samples
sinfo_umap <- cov_npx$sinfo %>% filter(Analysis_date == "2020-06-23", !grp %in% c("Neg Ctrl", "IPC"))
npx_umap <- cov_npx$npx[sinfo_umap$sample_id, ]
npx_umap <- npx_umap[complete.cases(npx_umap), ]
sinfo_umap <- sinfo_umap %>% filter(sample_id %in% rownames(npx_umap)) %>%
  # Update serostatus names
  mutate(serostatus = case_when(serostatus == "no reactivity" ~ "Seronegative, IgM-IgG-",
                                serostatus == "IgG + IgM" ~ "Seropositive, IgM+IgG+"))

sinfo_spk <- cov_npx$sinfo %>% filter(Analysis_date == "2020-11-13", !grp %in% c("Neg Ctrl", "IPC"), !str_detect(Analysis_group, "UMAP"))
npx_spk <- cov_npx$npx[sinfo_spk$sample_id, ]
npx_spk <- npx_spk[complete.cases(npx_spk), ]
sinfo_spk <- sinfo_spk %>% filter(sample_id %in% rownames(npx_spk)) %>%
  # Update serostatus names
  mutate(serostatus = case_when(serostatus == "IgG" ~ "Late phase, IgM-Ig+",
                                serostatus == "IgM" ~ "Early phase, IgM+IgG-"))

```

Also get dataset 3

```{r}

ds3 <- load_cov_npx(paste0("ds3_", cov_npx$data_set))
sinfo_ds3 <- ds3$sinfo %>% rename(serostatus = grp) # %>% filter(both_pos | !any_pos)
npx_ds3 <- ds3$npx %>% filter(sample_id %in% sinfo_ds3$sample_id)
sinfo_ds3 <- sinfo_ds3 %>% filter(sample_id %in% npx_ds3$sample_id) %>%
  # Update serostatus names
  mutate(serostatus = case_when(serostatus == "Negative" ~ "Seronegative, IgG-",
                                serostatus == "Positive" ~ "Seropositive, IgG+"))
binder_ds3 <- ds3$binfo

```

Loaded data set **`r ds3$data_set`** from **`r ds3$data_path`**

Load the median vs IQR outlier plots.

```{r}
outl_plts <- readRDS(paste0(out_dir, "../../data_dir/outliers_plt_obj_abspqn.rds"))
```


# Generate plots

```{r}
# Make plots per dataset
plts <- lapply(list(
  "Dataset1" = list("npx" = npx_umap, "sinfo" = sinfo_umap, "binder" = binder %>% rename(panel_long = panel, panel = panel_name), "name" = "Dataset1"),
  "Dataset2" = list("npx" = npx_spk, "sinfo" = sinfo_spk, "binder" = binder %>% rename(panel_long = panel, panel = panel_name), "name" = "Dataset2"),
  "Dataset3" = list("npx" = npx_ds3 %>% column_to_rownames("sample_id"),
                    "sinfo" = sinfo_ds3[match(npx_ds3$sample_id, sinfo_ds3$sample_id), ],
                    "binder" = binder_ds3 %>% rename(unique_id = assay),
                    "name" = "Dataset3")
), function(x) {
  
  # Make long data
  plt_dat <- x$npx %>% as.data.frame() %>% rownames_to_column("sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "unique_id", values_to = "npx") %>%
    left_join(x$binder %>% select(unique_id, panel), by = "unique_id") %>%
    left_join(x$sinfo %>% select(sample_id, serostatus), by = "sample_id") %>%
    mutate(serostatus = as.character(serostatus))
  
  # Choose different colours based on which dataset it is
  dat_col <- RColorBrewer::brewer.pal(6, "Set2")
  dat_col <- switch(x$name, "Dataset1" = dat_col[1:2], "Dataset2" = dat_col[3:4], "Dataset3" = dat_col[6:5]) %>%
    setNames(unique(plt_dat$serostatus))
  
  # Same point size for all
  point_size <- 3
  
  # PCA
  pca_plt <- plt_dat %>%
    # Perform per panel. To include variance explained in axes, make a plot per panel by nesting (dplyr) instead of facetting (ggplot)
    group_by(panel) %>% nest() %>%
    mutate(pca_plt = map(data, ~ {
      p <- .x %>% pivot_wider(id_cols = sample_id, names_from = "unique_id", values_from = "npx") %>%
        column_to_rownames("sample_id") %>%
        prcomp(., center = T, scale. = T) %>% summary()
      
      plt_dat <- data.frame(pc1 = p$x[, 1], pc2 = p$x[, 2],
                        sample_id = rownames(p$x), panel = panel) %>%
        # Add serostatus
        left_join(.x %>% select(sample_id, serostatus) %>% distinct(), by = "sample_id")
      
      # Data for drawing lines
      line_df <- plt_dat %>% filter(str_detect(sample_id, "_[:alpha:]$")) %>%
        mutate(line_id = str_remove(sample_id, "_[:alpha:]$")) %>%
        select(line_id, pc1, pc2)
      
        # Add serostatus
      plt<- ggplot(plt_dat, aes(x = pc1, y = pc2, colour = serostatus)) +
        geom_line(aes(group = line_id), colour = "grey", data = line_df) +
        geom_point(size = point_size) +
        labs(x = paste0("PC1 (", signif(p$importance[2, 1] * 100, 2), "% variance explained)"),
             y = paste0("PC2 (", signif(p$importance[2, 2] * 100, 2), "% variance explained)"),
             colour = "Serostatus") +
        scale_colour_manual(values = dat_col) +
        facet_wrap(~ panel) +
        theme_classic()
    })) %>%
    pull(pca_plt)
  
  # UMAP
  umap_plt <- plt_dat %>%
    # Perform per panel
    group_by(panel) %>% nest() %>%
    mutate(umap_plt = map(data, ~ {
      u <- .x %>% pivot_wider(id_cols = sample_id, names_from = "unique_id", values_from = "npx") %>%
        column_to_rownames("sample_id") %>%
        umap()
      
      plt_dat <- data.frame(UMAP1 = u$layout[, 1], UMAP2 = u$layout[, 2],
                        sample_id = rownames(u$layout), panel = panel) %>%
        # Add serostatus
        left_join(.x %>% select(sample_id, serostatus) %>% distinct(), by = "sample_id")
      
      # Data for drawing lines
      line_df <- plt_dat %>% filter(str_detect(sample_id, "_[:alpha:]$")) %>%
        mutate(line_id = str_remove(sample_id, "_[:alpha:]$")) %>%
        select(line_id, UMAP1, UMAP2)
      
      plt <- ggplot(plt_dat, aes(x = UMAP1, y = UMAP2, colour = serostatus)) +
        geom_line(aes(group = line_id), colour = "grey", data = line_df) +
        geom_point(size = point_size) +
        scale_colour_manual(values = dat_col) +
        labs(colour = "Serostatus") +
        facet_wrap(~ panel) +
        theme_classic()
    })) %>%
    pull(umap_plt)
  return(list("pca_plt" = pca_plt,
              "umap_plt" = umap_plt,
              "dataset" = x$name))
})

```

```{r}
# Wrap together plots
wrp_plts <- lapply(1:3, function(x) {
  wrap_plots(c(plts[[x]]$pca_plt, plts[[x]]$umap_plt, outl_plts[[x]]),
             guides = "collect", nrow = 3, ncol = 3) +
    plot_annotation(title = plts[[x]]$dataset, theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))
})
```

Combined PCA, UMAP, and median vs IQR plots (without labels) are saved as PDF files.

```{r}
# Save plots
for (i in seq_along(wrp_plts)) {
  ggsave(filename = paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_overview_plots_", cov_npx$data_set, "_", i, ".pdf"),
         plot = wrp_plts[[i]],
         width = 11, height = 10)
}

```


# Session information

```{r session information}
sessionInfo()
```
