---
title: "Covid19 DBS Olink analysis"
author: "Creator: Tea Dodig-Crnkovic. Contributors: Leo Dahl, Tea Dodig-Crnkovic"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    code_folding: hide
    toc_float:
      collapsed: false 

---
<!-- highlight: textmate -->

<!-- Edit TOC style (background color) -->
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #EAEAEA;
    color: black;}
</style>

<!-- Set background color for highlighted text -->
<style>
div.highlight { 
  background-color: #EAEAEA;  <!-- #39AC91 (turquoise) -->
  color: #FFFFEF;
  padding: 10px;
  border-radius: 2px 2px 2px 2px;}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE)
```


<style type="text/css">

h1.title {
  font-size: 22px;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
    font-size: 15px;
}
h3 { /* Header 3 */
  font-size: 12px;
}

</style>

***
# Description

<div class = "highlight">
Covid19 DBS Olink panels CVD3, MET, CAM.   
Here, the sample is from an individual that has donated blood over time --> longitudinal analysis.
  
 
</div>

***
# Initialization
```{r initialization, eval=T, warning=F}
# empty global environment
rm(list = ls(all.names = TRUE))
```

# Load packages
```{r packages, eval=T, warning=F, message=FALSE}
# load packages
library(tidyverse)
library(DT)
library(gridExtra)
library(reshape2) #melt()
library(clValid) #SOTA
library(writexl)
library(xlsx)
library(ggrepel)
library(patchwork)
```

***
# Load files
```{r pathways, eval=t}

source("load_cov_npx.R")
cov_npx <- load_cov_npx("abspqn_adj")

# information on longitudinal sampling dates
metadata <- read.delim("longitudinal/covid19_longitudinal.txt", stringsAsFactors = F)

# uniprot ids
uni_id <- read.xlsx2(file="../results/uniprot_to_ensg/uniprot_IDs.xlsx",
                    sheetIndex = 4)

# get the values from the serological assays
# Skip entry of sample form a different person
ag_dat <- metadata[-which(metadata$id == "other_person"), ]
rownames(ag_dat) <- ag_dat$id

```

Merge measurements from columns measuring the same antigen

```{r}
# MFI values of antigens of interest to use for analysis
ag_mfi <- ag_dat %>% 
  select(which(str_detect(colnames(.), "SPK|RBD|NCP"))) 

# The antigen data is of the character class and contains commas, remove those
ag_mfi <- apply(ag_mfi, 2, function(x) {
  new_mfi <- gsub(",", "", x) 
  class(new_mfi) <- "numeric"
  return(new_mfi)
}) %>% as.data.frame()

# Take means of antigens of same type (e.g. SPK_01, SPK_02, ...), keep as list to keep names
ag_mfi_mean_list <- apply(ag_mfi, 1, function(x) {
  
  # Go through the three antigens 
  ag <- lapply(c("SPK"="SPK", "RBD"="RBD", "NCP"="NCP"), function(y) {
    
    # Go through IgG and IgM
    ig <- lapply(c("IGG"="IGG", "IGM"="IGM"), function(z) {
      
      mean_out <- mean(x[which(str_detect(names(x), 
                                          paste0("^", y, ".*", z, "$")))])
      
    }) %>% as.data.frame()
    
  }) %>% as.data.frame()
  
})
ag_mfi_mean <- bind_rows(ag_mfi_mean_list)
rownames(ag_mfi_mean) <- make.names(names(ag_mfi_mean_list))

# Change names and make "duplicates" (with the same values) to incorporate into protein data
ag_mfi_mean <- rbind(
  ag_mfi_mean %>% `rownames<-`(paste0(rownames(ag_mfi_mean), "_A")),
  ag_mfi_mean %>% `rownames<-`(paste0(rownames(ag_mfi_mean), "_B"))
)

```


***
# Set input data
```{r data_selection, eval=t, results='hide'}

dat <- as.matrix(cov_npx$longitudinal$npx)
pat <- as.data.frame(cov_npx$longitudinal$sinfo)
prot <- as.data.frame(cov_npx$binder)
norm_info <- cov_npx$norm
rownames(pat) <- pat$sample_id
rownames(prot) <- prot$unique_id

# only read in longitudinal data

samples_to_include <- "Longitudinal sampling"
binders_to_include <- paste0("All binders included")

pat <- pat[match(rownames(dat),rownames(pat)),]
dim(dat); dim(pat)

# Include serology data
dat <- dat %>% as.data.frame() %>% rownames_to_column("samp") %>%
  left_join(ag_mfi_mean %>% rownames_to_column("samp"), by = "samp") %>%
  column_to_rownames("samp") %>% as.matrix()

prot_ags <- prot[1:6, ]
ags <- colnames(ag_mfi_mean)
rownames(prot_ags) <- ags
prot_ags[, c("assay", "id", "unique_id", "gene_name")] <- c(ags, ags, ags, ags)
prot <- rbind(prot, prot_ags)

```


***
# Configure output options
```{r output_options, eval=t}

# set output directory
out_dir <- "../results/longitudinal/"

# select samples and binders to include
selected_samples <- rownames(dat)
selected_binders <- colnames(dat)

samp_info <- paste0("Samples n=", dim(dat[selected_samples, selected_binders])[1])
binder_info <- paste0("Binders n=", dim(dat[selected_samples, selected_binders])[2])

```

***
**Data options**

*Data set*:   
`r cov_npx$data_set`

*Data set input file*:   
`r cov_npx$data_path`

*Data set notes*:   
`r norm_info`

*Samples included in analysis:*   
`r samples_to_include`

*Number of samples included:*   
`r samp_info`

*Binders included in analysis:*   
`r binders_to_include`

*Number of binders included:*   
`r binder_info`

***
# Add metadata
```{r eval=t, results='hide'}
metadata$unique_participant_ID <- make.names(metadata$id)

pat <- 
  metadata %>% 
  select(unique_participant_ID,
         Comment) %>% 
  rename("days_since_symptoms_onset"= "Comment") %>%  #"Comment" = days since onset of symptoms
  right_join(pat, by="unique_participant_ID") %>% 
  select(subj.id:sample_id, unique_participant_ID, days_since_symptoms_onset) %>% 
 mutate(rep = strsplit(sample_id, "_") %>% 
           lapply(`[[`, 2) %>% 
           unlist())
```


***
# Correlation between replicate binders
## Plot correlation
<details><summary>Click to show plots</summary>
```{r fig.align='center', warning=F}

# Get samples available in duplicates
dupl_samp <- pat %>% filter(unique_participant_ID %in% (pat %>% filter(duplicated(unique_participant_ID)) %>% pull(unique_participant_ID))) %>%
  pull(sample_id)

# put in donation order
dat_order <- dat[sort(dupl_samp), ]

par(mfrow=c(2,3), pty="s")

for(binder in sort(selected_binders)){

  x_min <- min(dat_order[grep("_A", rownames(dat_order)), binder],
    dat_order[grep("_B", rownames(dat_order)), binder], na.rm = T)
  x_max <- max(dat_order[grep("_A", rownames(dat_order)), binder],
    dat_order[grep("_B", rownames(dat_order)), binder], na.rm=T)
  
  plot(dat_order[grep("_A", rownames(dat_order)), binder],
    dat_order[grep("_B", rownames(dat_order)), binder],
    xlab="Rep A", ylab="Rep B", main=binder,
    xlim=c(x_min, x_max),
    ylim=c(x_min, x_max), pch=19)
  abline(0,1, col="blue")
  
  mtext(paste0("Rho = ", cor(dat_order[grep("_A", rownames(dat_order)), binder],
    dat_order[grep("_B", rownames(dat_order)), binder],
    method = "spearman") %>% round(2)), cex=0.7)
    
}


```
</details>


## Calculate correlation
Calculate correlation between binders, select proteins for downstream analysis.
```{r fig.align='center', warning=F}

z_score_all <-
  dat %>%
  scale() %>% 
  data.frame()

# put in donation order
z_score_dupl <- z_score_all[sort(dupl_samp), ]

for(binder in selected_binders){

  prot$rep_cor[prot$unique_id %in% binder] <-
    cor(z_score_dupl[grep("_A", rownames(z_score_dupl)), binder],
    z_score_dupl[grep("_B", rownames(z_score_dupl)), binder],
    method = "spearman") %>% round(2)
}

rho_cut_off <- "no cut-off" #0.75
selected_cor_binders <- selected_binders #prot$unique_id[which(prot$rep_cor >= 0.75)]
n_selected_cor_binders <- selected_cor_binders %>% length()

```
**Selected binders at Rho cut-off:**
`r rho_cut_off`

**N binders after cut-off:**
`r n_selected_cor_binders`


***
## Caluclate average z-score between replicates
Calculate average z-score per protein, for proteins above cut-off.
```{r fig.align='center', warning=F}

select_samps <- substr(rownames(z_score_all), 1, 12) %>% unique()

z_score_mean <- data.frame(matrix(nrow = length(select_samps),
                                  ncol = length(selected_binders)))

colnames(z_score_mean) <- selected_binders 
rownames(z_score_mean) <- select_samps

for (samp in select_samps) {
  z_score_mean[samp, ] <- apply(z_score_all[paste0(samp, c("_A", "_B")), ], 2, mean, na.rm=T)
}


# melt data (both replicates + mean)
df <- rbind(z_score_all, z_score_mean)

df <-
  df %>% 
  as_tibble() %>% 
  transform("unique_participant_ID" = substr(rownames(df), 1, 12),
            "sample_id" = rownames(df),
            "replicate" = substr(rownames(df), 14, 14))

levels(df$replicate) <- c("mean_A_B", "A", "B")

df$days_since_symptoms_onset <- pat$days_since_symptoms_onset[match(df$unique_participant_ID, pat$unique_participant_ID)]
  
df_melt <-
  df %>% 
  melt(c("sample_id", "unique_participant_ID", "days_since_symptoms_onset", "replicate")) %>%
  mutate(replicate = case_when(replicate == "" ~ "mean_A_B", T ~ replicate))


```


***
# Longitudinal plots
Plots per protein for longitudinal data.
Inlcude both replicates and average value.
```{r  eval=t, results='hide', message=FALSE, warning=F}

plot_list <- list()

for(binder in prot$unique_id[order(prot$rep_cor, decreasing = T)]) { #sort by correlation
  #binder <- selected_binders[1]

    plot_list[[binder]] <-
      df_melt %>%
  filter(variable == binder) %>%
  ggplot((aes(x = days_since_symptoms_onset,
              y = value,
              group = replicate,
              color = replicate))) +
  theme_light() +
  geom_point() +
  geom_line(aes(linetype=replicate)) +
  labs(title = paste0(binder, ", Rho = ", prot$rep_cor[prot$unique_id %in% binder]),
       subtitle = paste0(norm_info, ", ", cov_npx$data_set,
                         "\n", samples_to_include, ", ", binders_to_include),
       tag=format(Sys.time(),"%Y-%m-%d"),
       caption="")

}

spl <- split(plot_list, (seq_along(plot_list)-1) %/% 12)

ppl <- lapply(spl, function(g)
	marrangeGrob(grobs = g, layout_matrix = matrix(1:12, ncol=3, byrow=T), top=NULL))


pdf(paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"),
					 "_longitudinal_lineplots.pdf"),
		width = 14, height = 12, useDingbats=F, onefile=T)

print(ppl)

dev.off()

```

***
### Example plots
All plots can be found in output folder.
```{r fig.align='center', warning=F, fig.dim = c(14,12)}

print(ppl[1])

```


***
### Selected targets
Plot selected increasing and decrasing protein profiles.
```{r fig.align='center', warning=F, fig.dim = c(14,12)}
# red = decreasing, blue = increasing
red_tar <- c("ANGPTL7.MET", "IL.18BP.CVD3", "IL2.RA.CVD3") #"MBL2.CAM",  "NPDC1.MET"
blue_tar <- c("CHRDL2.MET", "CRTAC1.CAM", "Gal.3.CVD3") #"NADK.MET", "LYVE1.CAM"

df_melt$subgroup <- NA

df_melt$subgroup[df_melt$variable %in% c(red_tar)] <- "decreasing"
df_melt$subgroup[df_melt$variable %in% c(blue_tar)] <- "increasing"

# add gene names
df_melt <- 
  prot[, c("gene_name", "unique_id")] %>% 
  rename("variable"="unique_id") %>% 
  left_join(df_melt, by="variable")


pdf(paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"),
					 "_longitudinal_selected_lineplots.pdf"),
		width = 8, height = 5, useDingbats=F, onefile=T)

df_melt[df_melt$variable %in% c(red_tar, blue_tar),] %>% 
  filter(replicate == "mean_A_B") %>% 
  ggplot((aes(x = days_since_symptoms_onset,
              y = value,
              group = subgroup, #variable
              color = subgroup))) +
  theme_light() +
  #geom_line() +
  geom_smooth(method=loess, se=T) + #geom_text()
  geom_text_repel(aes(label=ifelse(days_since_symptoms_onset ==  33,as.character(gene_name),'')),hjust=0.7,vjust=0, size=2.5) + #color="black", 
  labs(title = paste0("Selected binders"),
       subtitle = paste0(norm_info, ", ", cov_npx$data_set,
                         "\n", samples_to_include, ", ", binders_to_include),
       tag=format(Sys.time(),"%Y-%m-%d"),
       caption="")

dev.off()
```


```{r, warning=F}
# Publication version

# Proteins to be included, constant, increasing or decreasing with time
const <-c("CPB1", "VWF", "ANGPTL1", "FCGR2A", "LILRB1", "CHIT1", "MBL2")
incr <- c("CCL15", "ANG", "CHRDL2", "CRTAC1", "EPCAM", "LGALS3") #, "FETUB", "IGFBP3")
decr <- c("GP1BA", "IL2RA", "PILRB", "CCL5", "IL18BP", "CHL1", "GNLY") #, "TGFBI", "CCL2", "IL1RL1")

df_melt$subgroup <- NA
df_melt$subgroup[df_melt$gene_name %in% const] <- "Constant"
df_melt$subgroup[df_melt$gene_name %in% incr] <- "Increasing"
df_melt$subgroup[df_melt$gene_name %in% decr] <- "Decreasing"

# Constant proteins
lineplt_const <- df_melt %>%
  filter(subgroup == "Constant" & replicate == "mean_A_B") %>%
  ggplot(aes(x = days_since_symptoms_onset,
             y = value)) +
  theme_classic() +
  geom_smooth(colour = "black", method = loess, se = T, size = 1.5) +
  labs(x = "Days since symptoms onset", y = "Z-score") +
  ylim(-2.5, 2.5) +
  annotate("text", x = 31, y = -1.3, fontface = "bold", size = 5.5,
           label = paste(const, collapse = "\n"), colour = "black") +
  geom_hline(yintercept = 0, size = 1, linetype = "dotted")

# Non-constant proteins
lineplt_nonconst <- df_melt %>%
  filter(subgroup %in% c("Increasing", "Decreasing") & replicate == "mean_A_B") %>%
  ggplot() +
  theme_classic() +
  geom_smooth(aes(x = days_since_symptoms_onset,
                  y = value,
                  group = subgroup,
                  colour = subgroup),
              method = loess, se = T, size = 1.5) +
  labs(x = "Days since symptoms onset", y = "Z-score") +
  annotate("text", x = 31, y = 1.7, fontface = "bold", size = 5.5,
           label = paste(incr, collapse = "\n"), colour = scales::hue_pal()(2)[2]) +
  annotate("text", x = 31, y = -0.3, fontface = "bold", size = 5.5,
           label = paste(decr[1:2], collapse = "\n"), colour = scales::hue_pal()(2)[1]) +
  annotate("text", x = 31, y = -1.75, fontface = "bold", size = 5.5,
           label = paste(decr[3:length(decr)], collapse = "\n"), colour = scales::hue_pal()(2)[1]) +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  ylim(-2.5, 2.5) +
  geom_hline(yintercept = 0, size = 1, linetype = "dotted")

# IgG/IgM measurements of SPK/RBD/NCP antigens
# MFI values of antigens of interest to use for analysis
ag_mfi <- metadata %>%
  filter(id != "other_person") %>%
  select(id, which(str_detect(colnames(.), "SPK|RBD|NCP")))
rownames(ag_mfi) <- make.names(ag_mfi$id)

# The antigen data is of the character class and contains commas, remove those
ag_mfi[, -1] <- apply(ag_mfi[, -1], 2, function(x) {
  new_mfi <- gsub(",", "", x) 
  class(new_mfi) <- "numeric"
  return(new_mfi)
}) %>% as.data.frame()

ag_mfi_melt <- ag_mfi[, -1] %>%
  as.data.frame() %>%
  mutate(id = rownames(ag_mfi)) %>%
  mutate(days_since_symptoms_onset = df[match(id, df$sample_id), "days_since_symptoms_onset"]) %>%
  pivot_longer(cols = -c(id, days_since_symptoms_onset)) %>%
  mutate(ig = str_extract(name, "IG(G|M){1}"))

lineplt_ig <- ggplot(ag_mfi_melt, aes(x = days_since_symptoms_onset, y = value, group = ig, linetype = ig)) +
  geom_smooth(method = loess, se = T, size = 1.5, colour = "black") +
  labs(x = "Days since symptoms onset", y = "MFI [AU]") +
  theme_classic() +
  theme(legend.position = "none") +
  scale_y_log10() +
  annotate("text", x = 32, y = 7500, fontface = "bold", size = 7,
           label = "IgG", colour = "black") +
  annotate("text", x = 32, y = 700, fontface = "bold", size = 7,
           label = "IgM", colour = "black")

# Combine
lineplt_combine <- lineplt_const + lineplt_nonconst + lineplt_ig + plot_annotation(tag_levels = "A") &
  theme(axis.title = element_text(size = 22),
        axis.text = element_text(size = 20),
        plot.tag = element_text(size = 30),
        axis.line = element_line(size = 1))

ggsave(paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"),
					 "_longitudinal_selected_lineplots_panels.pdf"),
       plot = lineplt_combine, width = 24, height = 8)
```


***
# Z-scoring and linear models
```{r calc, eval=t, results='hide'}

# use transformed data, z-scores
# use average value

df <- z_score_mean %>% t()
days_since_onset <- pat$days_since_symptoms_onset[match(colnames(df), pat$unique_participant_ID)]


lm_fit <- sapply(selected_binders, function(i) lm(df[i,] ~ days_since_onset), simplify= F)
intercept <- sapply(names(lm_fit), function(i) coef(summary(lm_fit[[i]]))[1,1])
slope <- sapply(names(lm_fit), function(i) coef(summary(lm_fit[[i]]))[2,1])

sum_residuals <- sapply(names(lm_fit), function(i)
  lm_fit[[i]] %>%
    residuals() %>%
    abs() %>%
    sum())

r_squared <- sapply(names(lm_fit), function(i) summary(lm_fit[[i]])$r.squared) %>%
  round(3)

#plot(df["SPON1.CVD3",])

lm_summary <-
  tibble(names(lm_fit), intercept, slope, sum_residuals, r_squared) %>%
  rename("unique_id" = "names(lm_fit)")

# calculate standard deviation for 1) intercept, 2) slope, 3) sum of residuals
  # if individuals is mean+/-cut-off(e.g. 3)xSD, set score = 1, otherwise score=0
  sd_cut_off <- 2

  cut_off_1 <- sapply(c("intercept", "slope", "sum_residuals"),
         function(x) mean(lm_summary[x] %>% unlist()) + sd(lm_summary[x] %>% unlist())*sd_cut_off)

   cut_off_2 <- sapply(c("intercept", "slope", "sum_residuals"),
         function(x) mean(lm_summary[x] %>% unlist) - sd(lm_summary[x] %>% unlist())*sd_cut_off)


lm_summary <-
  lm_summary %>%
add_column(intercept_score = 0,
                 slope_score = 0,
                 sum_residuals_score = 0) %>%
    transform(intercept_score = ifelse(intercept >= cut_off_1["intercept"] | intercept <= cut_off_2["intercept"], 1, intercept_score)) %>%
     transform(slope_score = ifelse(slope >= cut_off_1["slope"] | slope <= cut_off_2["slope"], 1, slope_score)) %>%
     transform(sum_residuals_score = ifelse(sum_residuals >= cut_off_1["sum_residuals"] | sum_residuals <= cut_off_2["sum_residuals"], 1, sum_residuals_score))


```

***
# Table of results
```{r fig.align='center', warning=F}

lm_summary %>%
  mutate(intercept = round(intercept, 3),
         slope = round(slope, 3),
         sum_residuals = round(sum_residuals, 3)) %>%
  select(-ends_with("score") ) %>%
  datatable(caption = "Longitudinal diversity, linear regression model output",
            rownames=F,
            options = list(columnDefs = list(list(className = 'dt-center', targets = "_all"))
                           ))

lm_summary %>%
  select(unique_id,
         intercept_score,
         slope_score,
         sum_residuals_score) %>%
  mutate(sum = apply(lm_summary[,c("intercept_score", "slope_score", "sum_residuals_score")], 1, sum)) %>%
  datatable(caption = "Longitudinal diversity, linear regression model output",
            rownames=F,
            options = list(columnDefs = list(list(className = 'dt-center', targets = "_all"))
                           ))

```

***
## Output excel
```{r warning=F}

# add uniprot ids to output list
uni_id <- left_join(uni_id %>% distinct(), prot[, c("assay", "unique_id", "panel_name")])

lm_summary %>% 
  right_join(uni_id, by="unique_id") %>% 
  mutate(intercept = round(intercept, 3),
         slope = round(slope, 3),
         sum_residuals = round(sum_residuals, 3)) %>%
  select(-ends_with("score") ) %>%
  select(assay, uniprot_ID, panel_name, everything(), -unique_id, -gene_name) %>% 
  write_xlsx(path=paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"), "_lm_model_per_protein.xlsx"))

```



***
# SOTA clustering
## SOTA plots
Aim is to test and find with 3, 6, or 9 clusters.
Here, z-scored data, mean of A and B, selected binders.
```{r fig.align='center', warning=F, fig.dim = c(10,7)}

cluster_n <- c(4, 5, 6, 9)

sotaCl_results <- list()

# N.B "gene" = number of proteins

for(i in c(cluster_n)-1){ #sota will use i+1

  sotaCl_results[[i+1]] <- sota(t(z_score_mean[, selected_cor_binders]),i)
  plot(sotaCl_results[[i+1]],
       main=paste0("SOTA clust n = ",i+1, "\n", norm_info))
}

```

## Selected SOTA plot for export
Select SOTA plot, make clean version and save to pdf
```{r warning=F, echo=F}

# Make clean ggplot of SOTA results
# sotaCl_results[[i+1]]$data # values of each unique protein profile
# sotaCl_results[[i+1]]$clust # which cluster each protein belongs to
# 
# sotaCl_results[[i+1]]$tree # average value in red for each cluster
# sotaCl_results[[i+1]]$totals # how many proteins there are per cluster ("genes" in SOTA plot)
# 

# select i
pick_clust = 9
  
plot_gg_sota <- data.frame("cluster"= sotaCl_results[[pick_clust]]$clust, sotaCl_results[[pick_clust]]$data)
# add values for red line
red_avg <- sotaCl_results[[pick_clust]]$tree[, -c(2:3)]
colnames(red_avg)[colnames(red_avg) %in% "ID"] <- "cluster"
plot_gg_sota <- rbind(plot_gg_sota, red_avg)

plot_gg_sota$unique_id <- rownames(plot_gg_sota)
plot_gg_sota$unique_id[plot_gg_sota$unique_id %in% c(1:9)] <- c("w_avg_line")


s_test <- 
  plot_gg_sota %>% 
  melt(c("cluster", "unique_id"))

# set names for headers
for(j in unique(s_test$cluster)){
  
  s_test$cluster[s_test$cluster %in% j] <- 
    paste0("cluster ", j, ", n=", sotaCl_results[[i+1]]$totals[j])
}


s_test$col_group <- "gray80"
s_test$col_group[s_test$unique_id %in% "w_avg_line"] <- "red"

s_test$days <- metadata[match(s_test$variable, metadata$unique_participant_ID), "Comment"]

pdf(paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"),
					 "_selected_sota_n_",pick_clust,"_clusters_discrX.pdf"),
		width = 7, height = 7, useDingbats=F, onefile=T)

ggplot(s_test,
       aes(x=variable, y=value, group=unique_id, color=col_group)) +
  #geom_point(alpha=.7) +
  geom_line(alpha=.7, size = 1) +
  scale_x_discrete(labels=as.character(metadata$Comment[2:6])) +
  xlab("Days since symptoms onset") +
  ylab("Z-score") +
  theme_classic() +
  theme(axis.line = element_line(size = 1),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 14)) +
  scale_color_identity() +
  facet_wrap(~ cluster, ncol = 3)

dev.off()

pdf(paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"),
					 "_selected_sota_n_",pick_clust,"_clusters_contX.pdf"),
		width = 7, height = 7, useDingbats=F, onefile=T)

ggplot(s_test,
       aes(x=days, y=value, group=unique_id, color=col_group)) +
  #geom_point(alpha=.7) +
  geom_line(alpha=.7, size = 1) +
  xlab("Days since symptoms onset") +
  ylab("Z-score") +
  theme_classic() +
  theme(axis.line = element_line(size = 1),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        strip.text = element_text(size = 14)) +
  scale_color_identity() +
  facet_wrap(~ cluster, ncol = 3)

dev.off()

```

***
## SOTA table output
```{r eval=t,  warning=F}

# # select cluster
# select_cluster_n <- 6
# 
# # extract which proteins belong to each cluster
# #identical(rownames(sotaCl_results[[9]][1]$data), prot$unique_id)

prot_sota <- prot[prot$unique_id %in% selected_cor_binders,]

#prot_sota$sota_cluster <- sotaCl_results[[select_cluster_n]][4]$clust

prot_sota <- cbind(prot_sota, 
          sapply(cluster_n, function(x) sotaCl_results[[x]][4]$clust))

colnames(prot_sota)[c(length(colnames(prot_sota))-length(cluster_n)+1):length(colnames(prot_sota))] <- paste0("sota_cluster_", cluster_n)

prot_sota %>%
  select(unique_id, assay, panel_name, starts_with("sota_cluster")) %>%
  datatable(caption = "SOTA clustering",
            rownames=F,
            options = list(columnDefs = list(list(className = 'dt-center', targets = "_all"))
                           ))

# # output SOTA clusters
prot_sota %>%
  select(unique_id, assay, uniprot_ID, panel_name, gene_name, starts_with("sota_cluster")) %>%
  arrange(unique_id) %>% 
  #arrange(sota_cluster) %>%
  write_xlsx(path=paste0(out_dir, format(Sys.time(),"%Y-%m-%d_%H%M%S"), "_sota_clusters_gene_names.xlsx"))

```


***
# Session information
```{r session_info, eval=T, echo=F}
sessionInfo()

```
