---
title: "Sample correlation plots"
author: "Leo Dahl"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    code_folding: hide
    toc_float:
      collapsed: false 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style type="text/css">

h1 { 
  font-size: 25px;
  margin-top: 1.5cm;
  margin-bottom: 0.5cm;
}
h2 { 
  font-size: 18px;
  margin-top: 1cm;
  margin-bottom: 0.5cm;
}

</style>

Description
===========

> This script makes heatmaps showing protein-protein correlation

```{r prep, message=F, results="hide"}

rm(list=ls())
options(stringsAsFactors = F)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(ggplotify)
library(patchwork)
library(cluster)
library(ggplot2)
library(stringr)
library(heatmaply)
library(dendextend)
library(openxlsx)

```

***

Load files
==========

```{r load data}

source("load_cov_npx.R")

cov_npx <- load_cov_npx("abspqn_adj")

source("sep_datasets.R")

```

Loaded data set **`r cov_npx$data_set`**  
from **`r cov_npx$data_path`**, with the following treatment:  
`r cov_npx$norm`

```{r}

npx <- cov_npx$npx
sinfo <- cov_npx$sinfo
binder <- cov_npx$binder

# Turn into character to not get unexpected behaviour from factor
sinfo$grp <- as.character(sinfo$grp)

# Merge duplicates/triplicates
sep_dat <- sep_datasets(npx, sinfo)
npx_umap <- sep_dat$npx_umap
npx_spk <- sep_dat$npx_spk
sinfo_umap <- sep_dat$sinfo_umap %>% mutate(serostatus = case_when(serostatus == "no reactivity" ~ "Seronegative, IgM-IgG-",
                                                                   serostatus == "IgG + IgM" ~ "Seropositive, IgM+IgG+"))
sinfo_spk <- sep_dat$sinfo_spk %>% mutate(serostatus = case_when(serostatus == "IgM" ~ "Early phase, IgM+IgG-",
                                                                 serostatus == "IgG" ~ "Late phase, IgM-IgG+"))
rm(sep_dat)

```

Also get dataset 3

```{r}

ds3 <- load_cov_npx(paste0("ds3_", cov_npx$data_set)) %>% merge_repl()

sinfo_ds3 <- ds3$sinfo %>% rename(serostatus = grp) %>%
  mutate(serostatus = case_when(serostatus == "Negative" ~ "Seronegative, IgG-",
                                serostatus == "Positive" ~ "Seropositive, IgG+"))
npx_ds3 <- ds3$npx %>% filter(sample_id %in% sinfo_ds3$sample_id)
binder_ds3 <- ds3$binfo

```

Loaded data set from **`r ds3$data_path`**

***

Protein-protein correlation
===========================

Protein-protein correlation heatmaps are made for the two different data sets. Interactive heatmaps below, static pdf versions and tables are found in results/correlation_plots. 

```{r clustering and heatmaps}
# Clustering function
cor_hclust <- function(x, k = NULL, clust_method = "ward.D") {
  # Cluster with correlation distance
  hcl <- as.dist(1 - x) %>%
    hclust(method = clust_method) %>%
    as.dendrogram() %>%
    ladderize()
  
  # Output, append cluster memberships if k is given
  out_list <- list()
  
  if (!is.null(k)) {
    # Cluster memberships
    cl <- cutree(hcl, k)
    out_list$cluster <- cl
  }
  
  out_list$dend <- hcl
  
  return(out_list)
}

# Optimal number of clusters according to gap statistic
get_optimal_k <- function(x, k_max = 15, b = 50) {
  gap_stat <- clusGap(x, FUNcluster = cor_hclust, K.max = k_max, B = b)
    
    # Get optimal k as smallest k where gap_k >= gap_k+1 - SE_k+1
    gap_tab <- gap_stat$Tab %>%
      as.data.frame() %>%
      # Gap minus SE for next k
      mutate(gms = lead(gap) - lead(SE.sim),
             gap_larger = gap >= gms)
    
    optimal_k <- which(gap_tab$gap == (gap_tab %>% filter(gap_larger) %>% pull(gap) %>% min()))
    
    return(optimal_k)
}

# Hierarchical clustering, heatmaps and alluvial plots
set.seed(12345)
cor_serostatus <- rbind(
  # Make long format data so that the datasets can be combined (do not necessarily contain the same proteins)
  npx_umap %>% as.data.frame() %>% rownames_to_column("unique_participant_ID") %>% pivot_longer(cols = -unique_participant_ID, names_to = "assay", values_to = "npx"),
  npx_spk %>% as.data.frame() %>% rownames_to_column("unique_participant_ID") %>% pivot_longer(cols = -unique_participant_ID, names_to = "assay", values_to = "npx"),
  npx_ds3 %>% as.data.frame() %>% rename(unique_participant_ID = sample_id) %>% pivot_longer(cols = -unique_participant_ID, names_to = "assay", values_to = "npx")
  ) %>%
  # Add info on serostatus
  left_join(
    rbind(sinfo_umap %>% select(unique_participant_ID, serostatus),
          sinfo_spk %>% select(unique_participant_ID, serostatus),
          sinfo_ds3 %>% rename(unique_participant_ID = sample_id) %>% select(unique_participant_ID, serostatus)),
    by = "unique_participant_ID"
  ) %>%
  mutate(serostatus = factor(serostatus, levels = c("Seronegative, IgM-IgG-",
                                                    "Seropositive, IgM+IgG+",
                                                    "Late phase, IgM-IgG+",
                                                    "Early phase, IgM+IgG-",
                                                    "Seropositive, IgG+",
                                                    "Seronegative, IgG-"))) %>%
  arrange(serostatus) %>%
  # Perform analysis per serostatus
  group_by(serostatus) %>%
  nest() %>%
  # Make wide format data per dataset
  mutate(data = map(data, ~ {
    pivot_wider(.x, id_cols = unique_participant_ID, names_from = "assay", values_from = "npx") %>%
      column_to_rownames("unique_participant_ID")
  })) %>%
  # Correlation matrix
  mutate(corr = map(data, cor, method = "spearman")) %>%
  # Optimal number of clusters using gap statistic
  mutate(optimal_k = map(corr, get_optimal_k, b = 50)) %>%
  # Heatmaps
  mutate(heatmap = map(corr, ~ {
    hcl <- cor_hclust(.x, k = optimal_k)
    clust <- hcl$dend
    clust_order <- hcl$cluster[order.dendrogram(clust)] %>% unique()
    
    hm <- Heatmap(.x, col = colorRamp2(breaks = seq(from = -1, to = 1, length.out = 101),
                                 colors = rev(colorRampPalette(brewer.pal(10, "RdBu"))(101))),
                  cluster_columns = clust, cluster_rows = clust, show_column_names = F, show_row_names = F,
                  name = "Rho", column_title = clust_order, column_split = as.numeric(optimal_k),
                  row_split = as.numeric(optimal_k), row_title = clust_order)
    # Add extra column title and grab expression to turn into ggplot for easier plot combining of plots
    hm <- draw(hm, column_title = as.character(serostatus), column_title_gp = gpar(fontsize = 18, fontface = "bold")) %>%
      grid.grabExpr() %>%
      as.ggplot()
  })) %>%
  # Cluster memberships
  mutate(mem = map(corr, ~ {cor_hclust(.x, k = optimal_k)$cluster})) %>%
  ungroup() %>%
  mutate(dataset = case_when(serostatus %in% c("Seronegative, IgM-IgG-", "Seropositive, IgM+IgG+") ~ "Dataset1",
                             serostatus %in% c("Late phase, IgM-IgG+", "Early phase, IgM+IgG-") ~ "Dataset2",
                             serostatus %in% c("Seropositive, IgG+", "Seronegative, IgG-") ~ "Dataset3")) %>%
  group_by(dataset) %>%
  nest() %>%
  # Merged heatmap per dataset
  mutate(heatmap_merge = map(data, ~ {
    wrap_plots(list(
      .x$heatmap[[1]] + labs(tag = switch(unique(dataset),
                                          "Dataset1" = "A",
                                          "Dataset2" = "B",
                                          "Dataset3" = "C")) + theme(plot.tag = element_text(size = 24)),
      .x$heatmap[[2]]
    ))
  })) %>%
  # Alluvial plot per dataset
  mutate(alv_plt = map(data, ~ {
    aclust <- cor_hclust(.x$corr[[1]], .x$optimal_k[[1]])$cluster
    bclust <- cor_hclust(.x$corr[[2]], .x$optimal_k[[2]])$cluster

    # Combine using protein order from a
    prots <- names(aclust)
    alluv_in <- data.frame(
      "protein" = prots,
      "clust_a" = paste0(as.character(.x$serostatus[[1]]), ": ", aclust),
      "clust_b" = paste0(as.character(.x$serostatus[[2]]), ": ", bclust[prots])
    )

    plt <- ggplot(alluv_in, aes(axis1 = clust_a, axis2 = clust_b)) +
      geom_alluvium() +
      geom_stratum() +
      geom_text(stat = "stratum",
                aes(label = after_stat(stratum))) +
      labs(tag = switch(unique(dataset),
                        "Dataset1" = "A",
                        "Dataset2" = "B",
                        "Dataset3" = "C")) +
      theme_void(14) +
      theme(plot.tag = element_text(size = 24))

    return(plt)
  }))

clust_mean_cor <- lapply(as.data.frame(t(expand.grid(1:2, 1:3))), function(x) {
    cor_mat <- cor_serostatus$data[[x[2]]]$corr[[x[1]]]
    clust <- cor_serostatus$data[[x[2]]]$mem[[x[1]]]
    clust_cor <- sapply(unique(clust), function(cl) {
        clust_prot <- names(clust[clust == cl])
        clust_cor <- cor_mat[clust_prot, clust_prot]
        # Keep only upper matrix
        clust_cor <- clust_cor[upper.tri(clust_cor, diag = F)]
        # Mean of the correlations
        mean(clust_cor)
    })
})
names(clust_mean_cor) <- lapply(as.data.frame(t(expand.grid(1:2, 1:3))), function(x) {
  cor_serostatus$data[[x[2]]]$serostatus[[x[1]]]
}) %>% unlist()
```


```{r save plots}
out_dir <- "../results/correlation_plots/"
# Save heatmaps
ggsave(filename = paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_heatmaps.pdf"),
       height = 18, width = 12, plot = wrap_plots(cor_serostatus$heatmap_merge, nrow = 3))

# Save alluvial plots
ggsave(filename = paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_alluvial.pdf"),
       height = 12, width = 8, plot = wrap_plots(cor_serostatus$alv_plt, nrow = 3))
```

```{r}
jaccard_ind <- function(a, b) {
  intersect_size <- length(intersect(a, b))
  union_size <-length(union(a, b))
  jacc_ind <- intersect_size / union_size
  
  return(jacc_ind)
}


mean_jaccard <- function(x, opt_k, mem, B = 50, samp_which = "both") {
  # Mean Jaccard Indices for clusters using bootstrapping. Hierarchical clustering with correlation distance
  # x, matrix to cluster
  # opt_k, numeric of length one specifiying the optimal number of clusters
  # mem, memberships from original clustering for which to compute mean jaccard indices
  # B, number of resamplings
  # samp_which, which to sample, "samples" or "proteins" or "both" (or any non-ambiguous starting substring)
  
  ji_vec <- sapply(1:B, function(iter) {
    if (grepl(paste0("^", samp_which), "samples", ignore.case = T) |
        grepl(paste0("^", samp_which), "both", ignore.case = T)) {
      
      # Get new samples, no need to remember which samples since the resulting protein-protein correlations will contain all proteins
      new_samps <- sample(1:nrow(x), nrow(x), replace = T)
      x <- x[new_samps, ]
      
    }
    
    if (grepl(paste0("^", samp_which), "proteins", ignore.case = T) |
        grepl(paste0("^", samp_which), "both", ignore.case = T)) {
      # Get a subset of proteins to cluster
      new_prots <- sample(colnames(x), ncol(x), replace = T)
      
      x <- x[, new_prots]
    }
    
    cor_mat <- cor(x, method = "spearman")
    
    # Cluster new correlation matrix
    new_mem <- cor_hclust(cor_mat, k = opt_k)$cluster
    
    # Use only proteins included in the new clustering to compute the index
    old_mem <- mem[unique(names(new_mem))]
    
    # Split names by cluster and compute Jaccard index between old and new clusters
    new_clust <- split(names(new_mem), new_mem)
    old_clust <- split(names(old_mem), old_mem)
    
    max_ji <- sapply(old_clust, function(a) {
      ji <- sapply(new_clust, function(b) {
        jaccard_ind(a, b)
      })
      
      return(max(ji))
    })
    
    return(max_ji)
  }) %>% t()
  
  out_df <- data.frame("cluster" = 1:opt_k,
                       "mean_ji" = colMeans(ji_vec),
                       "sd_ji" = apply(ji_vec, 2, sd))
  
  return(out_df)
}

set.seed(123)
mji <- lapply(as.data.frame(t(expand.grid(1:2, 1:3))), function(x) {
    in_dat <- cor_serostatus$data[[x[2]]]
    mean_jaccard(in_dat$data[[x[1]]],
                 in_dat$optimal_k[[x[1]]],
                 in_dat$mem[[x[1]]],
                 B = 50, samp_which = "prot")
})
names(mji) <- lapply(as.data.frame(t(expand.grid(1:2, 1:3))), function(x) {
  cor_serostatus$data[[x[2]]]$serostatus[[x[1]]]
}) %>% unlist()
```

```{r}
# Table of cluster assignments and mean jaccard indices
n_clust <- max(sapply(mji, nrow))
clust_tbl_in <- list(
  # Cluster assignments
  clust = data.frame("Binder" = colnames(cor_serostatus$data[[1]]$corr[[1]])) %>%
    # Additional binder information
    mutate(Gene_name = binder[match(Binder, binder$unique_id), "gene_name", drop = T],
           Panel = binder[match(Binder, binder$unique_id), "panel_name", drop = T],
           UniProt_ID = binder[match(Binder, binder$unique_id), "uniprot_ID", drop = T]),
  ji = data.frame("Cluster" = 1:max(sapply(mji, nrow))),
  mc = data.frame("Cluster" = 1:max(lengths(clust_mean_cor)))
)

for (grp in c("Seronegative, IgM-IgG-", "Seropositive, IgM+IgG+",
              "Early phase, IgM+IgG-", "Late phase, IgM-IgG+",
              "Seronegative, IgG-", "Seropositive, IgG+")) {
  # Dataset for matching position in cor_serostatus list
  ds <- case_when(grp %in% c("Seronegative, IgM-IgG-", "Seropositive, IgM+IgG+") ~ "Dataset1",
                  grp %in% c("Early phase, IgM+IgG-", "Late phase, IgM-IgG+") ~ "Dataset2",
                  grp %in% c("Seronegative, IgG-", "Seropositive, IgG+") ~ "Dataset3")
  
  # Add cluster memberships
  clust_mem_elem <- cor_serostatus$data[cor_serostatus$dataset == ds][[1]]
  clust_mem_ind <- str_which(clust_mem_elem$serostatus, regex(str_replace_all(grp, "\\+", "\\\\+"), ignore_case = T))
  clust_tbl_in$clust[, paste0("Cluster_", grp)] <- clust_mem_elem$mem[[clust_mem_ind]][clust_tbl_in$clust$Binder]
  
  # Add mean Jaccard indices
  mji_elem <- mji[[str_which(names(mji), regex(str_replace_all(grp, "\\+", "\\\\+"), ignore_case = T))]]
  clust_tbl_in$ji[, paste0(c("MJI_", "SD_"), grp)] <- c(
    c(mji_elem$mean_ji, rep(NA, n_clust - length(mji_elem$mean_ji))),
    c(mji_elem$sd_ji, rep(NA, n_clust - length(mji_elem$sd_ji)))
  )
  
  # Add mean cluster correlations
  mc_elem <- clust_mean_cor[[str_which(names(clust_mean_cor), regex(str_replace_all(grp, "\\+", "\\\\+"), ignore_case = T))]]
  clust_tbl_in$mc[, grp] <- c(mc_elem, rep(NA, n_clust - length(mc_elem)))
}
rm(clust_mem_elem, mji_elem, mc_elem)

clust_tbl <- createWorkbook()
addWorksheet(clust_tbl, "Cluster_assignments")
writeDataTable(clust_tbl, "Cluster_assignments", as.data.frame(clust_tbl_in$clust))
addWorksheet(clust_tbl, "Mean_Jaccard_Index")
writeDataTable(clust_tbl, "Mean_Jaccard_Index", clust_tbl_in$ji)
addWorksheet(clust_tbl, "Mean_Correlation_Rho")
writeDataTable(clust_tbl, "Mean_Correlation_Rho", clust_tbl_in$mc)

saveWorkbook(clust_tbl, paste0(out_dir, format(Sys.time(), "%Y-%m-%d_%H%M%S"), "_cluster_table.xlsx"))
```

```{r correlation table per serostatus}
# Make tables of protein correlation for the different data sets and analysis groups, here including p-values for the correlations
long_cor_mat <- function(x, respname = "rsp") {
  # Make long correlation matrix, skipping diagonals
  x[lower.tri(x, diag = T)] <- NA
  as.data.frame.table(x, responseName = respname) %>% na.omit()
}

cor_tbls <- sapply(cor_serostatus$dataset, function(ds) {
  # Get data for the dataset
  cor_serostatus_ds <- cor_serostatus$data[cor_serostatus$dataset == ds][[1]]
  
  # Go through serostatuses in the data
  sapply(as.character(cor_serostatus_ds$serostatus), function(ser) {
    d <- cor_serostatus_ds[cor_serostatus_ds$serostatus == ser, "data"][[1]][[1]]
    cor_test <- corr.test(d, method = "spearman", ci = F, adjust = "none")
    cor_mat <- long_cor_mat(cor_test$r, "Correlation")
    cor_p <- long_cor_mat(cor_test$p, "Pvalue")
    
    tbl_out <- cor_mat %>% left_join(cor_p, by = c("Var1", "Var2")) %>%
      mutate(FDR = p.adjust(Pvalue, method = "fdr")) %>%
      rename(Binder1 = Var1, Binder2 = Var2) %>%
      # Remove panel names from columns
      mutate(Binder1 = binder[match(Binder1, binder$unique_id), "assay", drop = T],
             Binder2 = binder[match(Binder2, binder$unique_id), "assay", drop = T])
    
  }, simplify = F)
  
}, simplify = F) %>%
  # Remove highest level (dataset) in list to get one element per serostatus
  unlist(recursive = F)

# Save tables
write.xlsx(cor_tbls, paste0( "../results/correlation_plots/", format(Sys.time(),"%Y-%m-%d_%H%M%S"), "_prot_cor_table.xlsx"))

```


***

Session information
===================

```{r session information}
sessionInfo()
```
